\proglang{Stan}: A Probabilistic Programming Language

### 2014-02-12 src/docs/papers/stan-language/article.tex
updated Stan paper and bib submission style for JStatSoft

%% an abstract and keywords
\Abstract{ \proglang{Stan} is a probabilistic programming language for
  specifying statistical models. A \proglang{Stan} program
  imperatively defines a log probability function over parameters
  conditioned on specified data and constants.  As of version 2.1,
  \proglang{Stan} provides full Bayesian inference for continuous-variable
  models through Markov chain Monte Carlo (MCMC) methods such as
  the No-U-Turn sampler (NUTS), an adaptive form of Hamiltonian Monte
  Carlo (HMC) sampling.  Penalized maximum likelihood estimates are
  calculated using optimization methods such as BFGS.

@@

  \proglang{Stan} is also a platform for computing log densities and
  their gradients and Hessians, which can be used in alternative
  algorithms such as variational Bayes, expectation propagation, and
  marginal inference using approximate integration.  To this end, Stan
  is set up so that the densities, gradients, and Hessians, along with
  intermediate quantities of the algorithm such as acceptance
  probabilities, are easily accessible.

  \proglang{Stan} can be called from the command line, through
  \proglang{R} using the \pkg{RStan} package, or through
  \proglang{Python} using the \pkg{PyStan} package.  All three
  interfaces support sampling or optimization-based inference and
  analysis, and \pkg{RStan} and \pkg{PyStan} also provide access to
  log probabilities, gradients, Hessians, and I/O transforms.
}

@@

We did not set out to build \proglang{Stan} as it currently exists.%
%
\footnote{\proglang{Stan}'s home page, \url{http://mc-stan.org/},
  links to the Stan manual, example models including translations of
  the BUGS examples and the models from \citep{GelmanHill:2007},
  getting started instructions and full documentation for Stan's
  interfaces for the command line shell (\pkg{CmdStan}), \proglang{Python}
  (\pkg{PyStan}), and \proglang{R} (\pkg{RStan}), and the source code repositories
  and issue trackers.}
%
Our original goal was to apply full Bayesian inference to the sort of multilevel
generalized linear models discussed in Part~II of
\citep{GelmanHill:2007}, which are structured with grouped and
interacted predictors at multiple levels, hierarchical covariance
priors, nonconjugate coefficient priors, latent effects as in
item-response models, and varying output link functions and
distributions.

@@

When we attempted to fit a time-series model, we found that
normalizing the data to unit sample mean and variance sped up the fits
by an order of magnitude.  In hindsight this isn't surprising, as the
performance of the numerical simulation in HMC scales with the
variation of the parameter scales.  In \proglang{Stan} 1.0 we
introduced an adaptive diagonal metric (mass matrix) into our HMC
implementations that allowed the parameter scales to be standardized
automatically; Stan 2.0 added an option to estimate a dense metric
(mass matrix) and a full covariance standardization of the posterior.

Because the diagonal and dense metrics perform only a global
standardization, the performance of \proglang{Stan} in models where
the relative parameter scales varies by location, such as hierarchical
and latent models, can suffer (hierarchical modeling is covered in
Sections~\ref{hierarchical-modeling.section} and reparameterizations to speed it
up in \ref{centering.section}).  Riemannian manifold Hamiltonian Monte
Carlo (RHMC) introduces a location-dependent metric that can overcome
these final hurdles \citep{GirolamiCalderhead:2011}; Stan has a
prototype implementation of RHMC based on the SoftAbs metric of
\citep{Betancourt:2012}, using a generalization of NUTS to Riemannian
manifolds \citep{Betancourt:2013}.

@@

Consider estimating the chance of success parameter for a Bernoulli
distribution based on a sequence of observed binary outcomes.  
Figure~\ref{bernoulli-model.fig} provides an implementation of such a
model in \proglang{Stan}.%
%
\footnote{This model is available in the \proglang{Stan} source
  distribution in \nolinkurl{src/models/basic\_estimators/bernoulli.stan}.}
%
\begin{figure}
\begin{Code}
data {
  int<lower=0> N;                  // N >= 0
  int<lower=0,upper=1> y[N];       // y[n] in { 0, 1 }
}
parameters {
  real<lower=0,upper=1> theta;     // theta in [0, 1]
}
model {
  theta ~ beta(1,1);               // prior
  y ~ bernoulli(theta);            // likelihood
}
\end{Code}
\caption{\it Model for estimating a Bernoulli parameter.}\label{bernoulli-model.fig}
\end{figure}
%
The model treats the observed binary data, \code{y[1],...,y[N]}, as
independent and identically distributed, with success probability
\code{theta}.  The vectorized likelihood statement can also be coded
using a loop as in BUGS, although it will run more slowly than the
vectorized form:
%
\begin{Code}
for (n in 1:N)
  y[n] ~ bernoulli(theta);
\end{Code}
%
A \code{beta(1,1)} (i.e., uniform) prior is placed on \code{theta},
although there is no special behavior for conjugate priors in
\proglang{Stan}.  The prior could be dropped from the model altogether
because parameters start with uniform distributions on their support,
here constrained to be between 0 and 1 in the parameter declaration
for \code{theta}.

@@

The output contiues with a CSV header naming the columns of the
output. For the default NUTS sampler in Stan 2.1.0, these are
%
\begin{Code}
lp__,accept_stat__,stepsize__,treedepth__,n_divergent__,theta
\end{Code}
%
The values for \code{lp\_\_} indicate the log probability (up to an
additive constant).  The column headed by \code{accept\_stat\_\_}
provides the Metropolis/slice acceptance statistic for each
iteration.%
%
\footnote{Acceptance is the usual notion for a Metropolis sampler such
  as HMC \citep{MetropolisEtAl:1953}.  For NUTS, the acceptance
  statistic is defined as the average acceptance probabilities of all
  possible samples in the proposed tree; NUTS itself uses a slice
  sampling algorithm for rejection \citep{Neal:2003, HoffmanGelman:2011}.}
%
The column \code{stepsize\_\_} indicates the step size
(i.e., time interval) of the simulated trajectory, while the column 
\code{treedepth\_\_} gives the tree depth for NUTS, defined as the 
log base 2 of the total number of steps in the trajectory.
The rest of the header will be the names of parameters; in this
example, \code{theta} is the only parameter.

@@

Stan follows the \proglang{C++} convention of separating words in
variable names using underbars (\code{\_}), rather than dots
(\code{.}), as used in \proglang{R} and \proglang{BUGS}, or camel case
as used in Java.

@@

\proglang{Stan} provides an especially rich set of special functions.
This includes all of the \proglang{C++} math library functions, as
well as numerous more specialized functions such as Bessel functions,
gamma and digamma functions, and generalized linear model link
functions and their inverses.  There are also many compound functions,
such as \code{log1m(x)}, which is more stable arithmetically for
values of x near 0 than \code{log(1 - x)}.  \proglang{Stan}'s special
functions are listed in Figures~\ref{special-functions.fig} and
\ref{special-functions-cont.fig}.

In addition to special functions, Stan includes distributions with
alternative parameterizations, such as \code{bernoulli_logit}, which
takes a parameter on the log odds (i.e., logit) scale.  This allows a
more concise notation for generalized linear models as well as more
efficient and arithmetically stable execution.

@@

Stan supports a growing collection of built-in univariate and
multivariate probability density and mass functions.  These
probability functions share various features of their declarations and
behavior.

@@

The list of probability functions is provided in Figures
\ref{prob-functions.fig}, \ref{prob-functions-cont.fig}, and
\ref{prob-functions-cont-2.fig}.

@@

In addition to performing full Bayesian inference via posterior
sampling, Stan also can perform optimization (i.e., computation of the
posterior mode).  We are currently working on implementing other
optimization-based inference approaches including variational Bayes,
expectation propagation, and and marginal inference using approximate
integration.  All these algorithms require optimization steps.

@@

Stan was and continues to be supported largely through grants from the
U.~S. government.  Grants which indirectly supported the initial
research and development included grants from the Department of Energy
(DE-SC0002099), the National Science Foundation (ATM-0934516), and the
Department of Education Institute of Education Sciences
(ED-GRANTS-032309-005 and R305D090006-09A).  The high-performance
computing facility on which we ran evaluations was made possible
through a grant from the National Institutes of Health
(1G20RR030893-01).  Stan is currently supported in part by a grant
from the National Science Foundation (CNS-1205516).

>>>>

%% an abstract and keywords
\Abstract{ \proglang{Stan} is a probabilistic programming language for
  specifying statistical models. A \proglang{Stan} program
  imperatively defines a log probability function over parameters
  conditioned on specified data and constants.  As of version 2.1,
  \proglang{Stan} provides full Bayesian inference for continuous-variable
  models through Markov chain Monte Carlo methods such as
  the No-U-Turn sampler, an adaptive form of Hamiltonian Monte
  Carlo sampling.  Penalized maximum likelihood estimates are
  calculated using optimization methods such as the
  Broyden–Fletcher–Goldfarb–Shanno algorithm.

@@

  \proglang{Stan} is also a platform for computing log densities and
  their gradients and Hessians, which can be used in alternative
  algorithms such as variational Bayes, expectation propagation, and
  marginal inference using approximate integration.  To this end, \proglang{Stan}
  is set up so that the densities, gradients, and Hessians, along with
  intermediate quantities of the algorithm such as acceptance
  probabilities, are easily accessible.

  \proglang{Stan} can be called from the command line, through
  \proglang{R} using the \pkg{RStan} package, or through
<  \proglang{Python} using the \pkg{PyStan} package.  All three
  interfaces support sampling or optimization-based inference and
  analysis, and \pkg{RStan} and \pkg{PyStan} also provide access to
  log probabilities, gradients, Hessians, and I/O transforms.
}

@@

We did not set out to build \proglang{Stan} as it currently exists.%
%
\footnote{\proglang{Stan}'s home page, \url{http://mc-stan.org/},
  links to the \proglang{Stan} manual, example models including translations of
  the \proglang{BUGS} examples and the models from \citep{GelmanHill:2007},
  getting started instructions and full documentation for \proglang{Stan}'s
  interfaces for the command line shell (\pkg{CmdStan}), \proglang{Python}
  (\pkg{PyStan}), and \proglang{R} (\pkg{RStan}), and the source code repositories
  and issue trackers.}
%
Our original goal was to apply full Bayesian inference to the sort of multilevel
generalized linear models discussed in Part~II of
\citep{GelmanHill:2007}, which are structured with grouped and
interacted predictors at multiple levels, hierarchical covariance
priors, nonconjugate coefficient priors, latent effects as in
item-response models, and varying output link functions and
distributions.

@@

When we attempted to fit a time-series model, we found that
normalizing the data to unit sample mean and variance sped up the fits
by an order of magnitude.  In hindsight this isn't surprising, as the
performance of the numerical simulation in HMC scales with the
variation of the parameter scales.  In \proglang{Stan} 1.0 we
introduced an adaptive diagonal metric (mass matrix) into our HMC
implementations that allowed the parameter scales to be standardized
automatically; \proglang{Stan} 2.0 added an option to estimate a dense metric
(mass matrix) and a full covariance standardization of the posterior.

Because the diagonal and dense metrics perform only a global
standardization, the performance of \proglang{Stan} in models where
the relative parameter scales varies by location, such as hierarchical
and latent models, can suffer (hierarchical modeling is covered in
Section~\ref{hierarchical-modeling.section} and reparameterizations to speed it
up in Section~\ref{centering.section}).  Riemannian manifold Hamiltonian Monte
Carlo (RHMC) introduces a location-dependent metric that can overcome
these final hurdles \citep{GirolamiCalderhead:2011}; \proglang{Stan} has a
prototype implementation of RHMC based on the SoftAbs metric of
\citep{Betancourt:2012}, using a generalization of NUTS to Riemannian
manifolds \citep{Betancourt:2013}.

@@

Consider estimating the chance of success parameter for a Bernoulli
distribution based on a sequence of observed binary outcomes.  
Figure~\ref{bernoulli-model.fig} provides an implementation of such a
model in \proglang{Stan}.%
%
\footnote{This model is available in the \proglang{Stan} source
  distribution in \nolinkurl{src/models/basic\_estimators/bernoulli.stan}.}
%
\begin{figure}
\begin{Code}
data {
  int<lower=0> N;                  // N >= 0
  int<lower=0,upper=1> y[N];       // y[n] in { 0, 1 }
}
parameters {
  real<lower=0,upper=1> theta;     // theta in [0, 1]
}
model {
  theta ~ beta(1,1);               // prior
  y ~ bernoulli(theta);            // likelihood
}
\end{Code}
\caption{\it Model for estimating a Bernoulli parameter.}\label{bernoulli-model.fig}
\end{figure}
%
The model treats the observed binary data, \code{y[1],...,y[N]}, as
independent and identically distributed, with success probability
\code{theta}.  The vectorized likelihood statement can also be coded
using a loop as in \proglang{BUGS}, although it will run more slowly than the
vectorized form:
%
\begin{Code}
for (n in 1:N)
  y[n] ~ bernoulli(theta);
\end{Code}
%
A \code{beta(1,1)} (i.e., uniform) prior is placed on \code{theta},
although there is no special behavior for conjugate priors in
\proglang{Stan}.  The prior could be dropped from the model altogether
because parameters start with uniform distributions on their support,
here constrained to be between 0 and 1 in the parameter declaration
for \code{theta}.

@@

The output contiues with a CSV header naming the columns of the
output. For the default NUTS sampler in \proglang{Stan} 2.1.0, these are
%
\begin{Code}
lp__,accept_stat__,stepsize__,treedepth__,n_divergent__,theta
\end{Code}
%
The values for \code{lp\_\_} indicate the log probability (up to an
additive constant).  The column headed by \code{accept\_stat\_\_}
provides the Metropolis/slice acceptance statistic for each
iteration.%
%
\footnote{Acceptance is the usual notion for a Metropolis sampler such
  as HMC \citep{MetropolisEtAl:1953}.  For NUTS, the acceptance
  statistic is defined as the average acceptance probabilities of all
  possible samples in the proposed tree; NUTS itself uses a slice
  sampling algorithm for rejection \citep{Neal:2003, HoffmanGelman:2011}.}
%
The column \code{stepsize\_\_} indicates the step size
(i.e., time interval) of the simulated trajectory, while the column 
\code{treedepth\_\_} gives the tree depth for NUTS, defined as the 
log base 2 of the total number of steps in the trajectory.
The rest of the header will be the names of parameters; in this
example, \code{theta} is the only parameter.

@@

\proglang{Stan} follows the \proglang{C++} convention of separating words in
variable names using underbars (\code{\_}), rather than dots
(\code{.}), as used in \proglang{R} and \proglang{BUGS}, or camel case
as used in Java.

@@

\proglang{Stan} provides an especially rich set of special functions.
This includes all of the \proglang{C++} math library functions, as
well as numerous more specialized functions such as Bessel functions,
gamma and digamma functions, and generalized linear model link
functions and their inverses.  There are also many compound functions,
such as \code{log1m(x)}, which is more stable arithmetically for
values of x near 0 than \code{log(1 - x)}.  \proglang{Stan}'s special
functions are listed in Figure~\ref{special-functions.fig} and
Figure~\ref{special-functions-cont.fig}.

In addition to special functions, \proglang{Stan} includes distributions with
alternative parameterizations, such as \code{bernoulli_logit}, which
takes a parameter on the log odds (i.e., logit) scale.  This allows a
more concise notation for generalized linear models as well as more
efficient and arithmetically stable execution.

@@

\proglang{Stan} supports a growing collection of built-in univariate and
multivariate probability density and mass functions.  These
probability functions share various features of their declarations and
behavior.

@@

The list of probability functions is provided in
Figure~\ref{prob-functions.fig}, 
Figure~\ref{prob-functions-cont.fig}, and
Figure~\ref{prob-functions-cont-2.fig}.

@@

In addition to performing full Bayesian inference via posterior
sampling, \proglang{Stan} also can perform optimization (i.e., computation of the
posterior mode).  We are currently working on implementing other
optimization-based inference approaches including variational Bayes,
expectation propagation, and and marginal inference using approximate
integration.  All these algorithms require optimization steps.

@@

\proglang{Stan} was and continues to be supported largely through grants from the
U.~S. government.  Grants which indirectly supported the initial
research and development included grants from the Department of Energy
(DE-SC0002099), the National Science Foundation (ATM-0934516), and the
Department of Education Institute of Education Sciences
(ED-GRANTS-032309-005 and R305D090006-09A).  The high-performance
computing facility on which we ran evaluations was made possible
through a grant from the National Institutes of Health
(1G20RR030893-01).  \proglang{Stan} is currently supported in part by a grant
from the National Science Foundation (CNS-1205516).
